---
title: "Data wrangling: more helpful functions"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Using paste, separate, and unite to tidy data

A common part of data wrangling is combining or splitting variables for use in analysis. Two powerhouse functions for this purpose are separate and unite. As their names, suggest, the first splits columns and the second combines them.

Let's start with the Waffle House dataset from Safegraph

```{r}
whdata<-read_csv("data/whdata_dec21_mar22.csv")
```

Right now, the dataset includes the address of each store in one column. However, sometimes we want to separate the house number from the name of the street. To do so, we need a separator. This can include some kind of punctuation (e.g., comma, dash, period), but even just a space can do. Here's how you could use that with separate:

```{r}
whdata_1<-whdata %>%
  separate(street_address,into=c("house_num","street"),sep=" ")
```

Looking at this data file allows you to see that this function worked, but you also got an error saying that much of the address was discarded! That's because there's more than one space. The `extra` parameter allows us to select what to do with that extra data. In this case, we'll merge it with everything else in the street column.

```{r}
whdata_2<-whdata %>%
  separate(street_address,into=c("house_num","street"),sep=" ",extra="merge")
```

Figuring out how to further parse this address--pulling out the road type (St., Rd., etc.)--would be much trickier, but separate along can help a lot with parsing text.

To *unite* data, we can simply list the variables we want to combine together. So to join address, city, and state, we could do the following. This code creates a new variable called `fulladd` that unites the three address variables.

```{r}
whdata_3<-whdata %>%
  unite(fulladd,c(street_address,city,region),sep = ", ")
```

Note that this removes all the component columns by default. We can use the `remove` parameter to keep them.

```{r}
whdata_4<-whdata %>%
  unite(fulladd,c(street_address,city,region),sep = ", ",remove=FALSE)
```

Lastly, paste is an all purpose function similar to unite that's used to concatenate data. Using paste allows you more flexibility in how the columns are combined. See this code for an example that creates a text description of each store's location

```{r}
whdata_5<-whdata %>%
  mutate(descript=paste(location_name," is located at ",street_address,
                        " which is in the city of ",city,", ",region,".",sep=""))
```

#####You try it!
The `bucketed_dwell_times` variable lists how many minutes people stayed at Waffle House. Use separate to divide those times into distinct columns.

###Further text cleaning

Sometimes in data wrangling you have to clean up text after you parse/organize it. For example, the "visits_by_day" field provides a list of the number of registered visits for each day of the weekly time period. What if we wanted to separate those into separate columns? We could use this code:

```{r}
whdata_6<-whdata %>%
  separate(visits_by_day,into=c("D1","D2","D3","D4","D5","D6","D7"),sep=",") 
```

Inspecting the data shows that `D1` and `D7` are problematic, though--they have square brackets from the original data field. We can use the `str_detect` function from the tidyverse (technically the `stringr` package) to erase those. The code below uses mutate, replacing the square brackets with a null character. Because those brackets are used in coding, we have to tell R that we mean the character version by adding two backslashes beforehand: `\\[` rather than just `]`.

```{r}
whdata_6<-whdata %>%
  separate(visits_by_day,into=c("D1","D2","D3","D4","D5","D6","D7"),sep=",") %>%
  mutate(D1=str_replace(D1,'\\[',''),
         D7=str_replace(D7,'\\]',''))
```

If you inspect the variables in this dataset, you'll see that those visit count columns (D1-D7) are currently characters, which makes sense since the original data was a character string. We can use the `mutate_at` function to transform those to numeric, as shown below:

```{r}
whdata_6a<-whdata_6 %>%
  mutate_at(c('D1','D2','D3','D4','D5','D6','D7'),as.numeric)
```

#####You try it!
Do more work with the bucketed_dwell_times variable to tidy those data. More specifically, you should have variable columns named after the dwell time categories (<5, 5-10, 11-20, etc.) and the count of users for each category listed under each one. Remove any excess characters that aren't meaningful.

If you want a further challenge, try to identify what other chains Waffle House patrons most commonly visit on the same day (`related_same_day_brand`)