---
title: 'Geog 4/6300: Lab 3'
output:
  github_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(sf)
```

## Spatial statistics and visualization

**Due:** Wednesday, Oct. 3

**Value:** 30 points

**Overview:**
This lab has three main sections: mapping point patterns, calculating rates and location quotients, and quadrat analysis. Your answers to the lab questions should be typed in the relevant spaces below and pushed to your Github repo. 

###Part 1: mapping the data and its distribution###

The lab folder on Google Drive has three files related to crime data for Spokane, Washington.  

* “Spokane_crimes_all” has data for every crime committed in from 2000-2015. 

* "SpokanePrecincts_data" has precinct boundaries, crime counts, the 2010 population of blocks in those precincts. Note: this is saved as a geojson file, which similar to a shapefile but more compact.

You can load each file using the commands below, using st_as_sf to convert the crimes to spatial data

**Question 1 (3 points)** _Load the Spokane_crimes_all.csv file in the lab3_data folder of this repo. Use st_as_sf to convert this data frame to spatial data (see script 6-3 for an example). Lastly, use st_read to import the precinct boundaries, which are in the spokanprecincts_data.geojson file. Call the Spokane_crimes_all file using kable when you are done._

```{r}
#Code goes here
```


**Question 2 (3 points):** _Filter the crime reports so that it contains just murder as the offense type. Using tmap, create a map that shows the location of all reports in the this data frame and the precinct boundaries. Also include a scale bar and north arrow. See script 6-3 for a model to work from._

```{r}
#Code goes here
```


**Question 3 (2 points):** _Create a summary table showing counts of each type of offense in these data using group_by and summarise. Use kable to show that table when done._

```{r}
#Code goes here
```


**Question 4 (3 points):** _Create a new data frame with the mean centers for each type of offense. This will work best if you start with the original data frame, the one without a geometry column. Then use ggplot and geom_point to plot these points on a graph, changing the color based on the offense._

```{r}
#Code goes here
```

**Question 5 (2 points)** _Interpret how geographic differences between the mean centers you calculated are meaningful. What do they tell you about the locations of these crimes?_

{Your answer here.}

**Question 6 (2 points):** _The nearest neighbor index for vehicle thefts in this dataset is 0.44 compared to 1.51 for murders. Explain what these two numbers tell you about the distribution of these two point datasets._

{Your answer here.}

###Part 2: Crime rates###
The precinct data frame has the number of crimes in each precinct. Crimes are commonly reported as crimes/100,000, so you should calculate the rate of any of these crimes rather than showing the raw counts. For this lab, we will focus on reported *burglary*. Calculate the assault rate by creating a new variable in the data, dividing the reported crimes by population and then multiplying by 100,000:

[new variable] <- [Burglary variable] / [population variable] * 100000

Use the mutate command from the tidyverse to create this variable.

**Question 7 (2 points)** _Select just the precinct number and burglary rate from the precincts data frame. Then create a new variable for burglary rate using the formula above. Create a histogram using ggplot showing the distribution of this new variable._ 

```{r}
#Code goes here
```

**Question 8 (3 points)** _Create a map of the burglary rate using tmap. Use style="jenks" in the tm_polygons function to use natural breaks classifciation. Also include the precinct number with the tm_text function. What's one geographic trend you see on this map?_

```{r}
#Code goes here
```

{Explanation goes here.}

**Question 9 (3 points)** _The burglary rate for the city of Spokane in this dataset is 10,639 per 100,000 residents. Using this figure, calculate the location quotient for burglary in each precinct within the city. Which precincts have the highest and lowest location quotients? What does this tell you about each of those precincts?_



###Part 3: Quadrat analysis###
You can also use the spatstat package to create a ppp object for the data on burglaries.

**Question 10 (2 points)** _Create a new data frame with just reported burglaries. Then create a ppp version of these data in the spatstat package using the methods shown in class. Plot this object to show the points._ 

**Question 11 (2 points)** _Create a kernel density map and appropriately sized quadrat grid for burglaries in Spokane during the study period using spatstat. You’ll need to include “add=TRUE” for the quadratcount function to layer it and change the text color to white as we did in class._

**Question 12 (3 points)** _In this lab, you have used mean center, calculated rates, location quotients, kernel density mapping, and quadrat analysis to analyze the distribution of crime in Spokane. Pick **two** of these and compare what they tell you about this dataset. What do you learn about these data from each?_ 
