---
title: 'Lab 2: Your responses'
output:
  pdf_document: default
  github_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Your name: {Name here}

In the spaces below, insert your code and/or explanation as notes. Look for the angle brackets-{like these}-for where a response is needed.

When you are done, click on the arrow next to "knit" and choose "Knit as PDF" to save. 

Load the data and the tidyverse package in the code chunk below

```{r data andlibrary}
library(tidyverse)
library(sf)
spokane_crimes_all<-read_csv("https://github.com/jshannon75/geog4300/raw/master/Labs/Lab%202_%20Point%20patterns%20and%20quadrat%20analysis/Spokane_crimes_all.csv")

spokane_crimes_sf<-st_as_sf(spokane_crimes_all,coords=c("Long","Lat"),crs=4326)

spokaneprecincts_data<-st_read("https://github.com/jshannon75/geog4300/raw/master/Labs/Lab%202_%20Point%20patterns%20and%20quadrat%20analysis/SpokanePrecincts_data.geojson")
```

**Question 1 (4 points):** _Using the Offense variable and working with the spokane_crimes_all data, select just Murder and Vehicle Theft as crimes. Using tmap, create a map that shows the location of all reports in the this data frame and the precinct boundaries. See script 6-2 for a model to work from or the example for question 3 below._

```{r question_1a}
#Create a data frame with just reports of murder and vehicle theft
```

```{r question_1b}
#Create your map of just murders
```

**Question 2 (2 points):** _Create a new data frame with the mean centers for each type of crime listed in the spokane_crime_sf data frame._ 

```{r question2}
#Create your data frame here

```

**Question 3 (3 points):** _Map the mean centers you created in question 2 using tmap. To do so, you'll need to use st_as_sf to convert your new data frame to sf format. Use tm_dots to show the points in different colors, as in the example above (though here you're showing only the mean centers)._

```{r question3}
#Code for creating the map of mean centers here

```

**Question 4 (2 points)** _Interpret how geographic differences between the mean centers you calculated are meaningful. What do they tell you about the locations of these crimes?_

{Your response here}

**Question 5 (2 points):** _The nearest neighbor index for vehicle thefts in this dataset is 0.44 compared to 1.51 for murders. Explain what these two numbers tell you about the distribution of these two point datasets._

{Your response here}

**Question 6 (5 points)** _Select just the precinct number, population (totpop), and burglary rate from the precincts data frame. Then create a new variable for this rate using the formula above. Then do the following: 1) Call your dataset once you've created it to show your results. 2) Create a histogram using ggplot showing the distribution of these data. 3) Map the rates using tmap._

```{r question 6a}
#Subset your data and calculate burglary rates. Call the data frame when done

```

```{r question 6b}
#Histogram of the burglary rates using ggplot

```

```{r question 6a}
#Map of the burglary rates using tmap

```

**Question 7 (2 points)** _The burglary rate for the city of Spokane in this dataset is 10,639 per 100,000 residents. Using this figure, calculate the location quotient for burglary in each precinct within the city._

```{r question 7}
#Calculate lQ

```

**Question 8 (3 points)** _Using the script from week 7 in class as a guide, combine the kernel density map and appropriately sized quadrat grid for robberies in Spokane during the study period. You’ll need to include “add=TRUE” for the quadratcount function to layer it and change the text color to white as we did in class._

```{r quetsion 8}
#Kernel density and quadrat map here

```

**Question 9 (4 points)** _Based on your analysis in the questions 6 through 8, describe the statistical and geographic distribution of burglary in the city of Spokane in a paragraph. Specifically address any skew in this data and the geographic location of the highest and lowest values for the burglary rate. Which of these visualizations was most helpful to you and why?_

{Your response here}

**Question 10 (2 points):** _Calculate the probability of observing each possible count of persons in a camping party for each reservation (including no shows)._ 

```{r }
#Calculate the probabilities here

```

**Question 11 (2 points):** _Calculate the probabilities the stream flow in a given year will exceed 42 and 48 CFM._


```{r }
#Calculate the probabilities here

```