---
title: "Predicting Waffle Houses (multivariate regression)"
output: html_notebook
editor_options: 
  chunk_output_type: console
---

Let's use multivariate regression to predict Waffle House density by county.

Loading packages and our datasets

```{r}
library(tidyverse)
library(sf)
library(corrplot)
library(car)

wh_cty<-read_csv("data/wh_locations_coords.csv") %>%
  count(cty_fips,name="wh_count") #Creates count of WH per county
census<-read_csv("data/ACSCtyData_2022ACS.csv") %>%
  left_join(wh_cty) %>%
  mutate(wh_100k=wh_count/totpop_race*100000)
```

What's the normality of our main variables?

```{r}
hist(census$wh_100k)
hist(census$pov_pop_pct)
hist(census$LessHS_pct)

```

What about multicollinearity? How are these variables related?

```{r}
census_sel<-census %>%
  filter(is.na(wh_100k)==FALSE) #Removes counties without Waffle Houses

correlation_matrix <- cor(census_sel %>% select(totpop_race,wht_pop_pct:fb_ins_pct,wh_count,wh_100k))
corrplot(correlation_matrix,order="hclust")

```

Here's a simple model:

```{r}
model<-lm(wh_100k~pov_pop_pct+LessHS_pct, data=census_sel)
summary(model)
```

Check for multicollinearity in the model using VIF:

```{r}
vif(model)
```

Look at normality of residuals.

```{r}
hist(residuals(model))
```

Model with a few more variables

```{r}
model1<-(lm(wh_100k~pov_pop_pct+fb_pct+wht_pop_pct+intnet_all_pct, 
           data=census_sel))
summary(model1)
vif(model1)
hist(residuals(model1))

```

What if we add States as an additional categorical variable?

```{r}
model2<-(lm(wh_100k~pov_pop_pct+fb_pct+wht_pop_pct+intnet_all_pct+st_name, 
           data=census_sel))
summary(model2)
```


When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.
