---
title: 'Geog 4/6300: Lab 3'
output:
  github_document: default
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Confidence Intervals and Sampling

**Due:** Monday, Oct. 16

**Value:** 30 points

**Overview:**
This lab covers  two main topics: basic spatial statistics and probability distributions. We will be using individual level “microdata” from the Current Population Survey (CPS). It is designed as an ongoing (collected monthly) set of data on financial and demographic characteristics. One main use of the CPS is to calculate national levels of food insecurity. Each December, a food security supplement is added to the regular survey, and data from the supplement is included here.

To load these data, load the csv file:

```{r message=FALSE}
library(tidyverse)
cps_data<-read_csv("https://github.com/jshannon75/geog4300/raw/master/Labs/Lab%203-Confidence%20intervals%20Sampling/IPUMS_CPS_FoodSec.csv")
```

This contains a csv file with microdata from the CPS that is de-identified and publically available through the Minnesota Population Center (https://cps.ipums.org/cps/index.shtml). There is also a [codebook available on Github](https://github.com/jshannon75/geog4300/raw/master/Labs/Lab%203-Confidence%20intervals%20Sampling/IPUMS_CPS_Codebook.pdf) describing each of those variables.

###Part 1: Calculating national food insecurity###
The n() function can be used to count records in a group. See this example for immigration:

```{r}
cps_data %>%
  group_by(YRIMMIG) %>%
  summarise(count=n())
```

For this response, look at the FSSTATUS variable, which describes the food security of respondents. While food security status is often grouped into “low” and “very low” food security, these two are often just combined to a single measure: food insecure. 

**Question 1 (4 points)** _Using the information on this variable in the codebook (p. 12), create a subset of records without the NIU or missing response records. Then use group_by and summarise to calculate the number of individuals grouped in each status as shown above.  Use the resulting data frame to calculate an estimate of the national food insecurity rate._

**Question 2 (3 points)** _Using the formula for confidence intervals for proportions shown in class, calculate a confidence interval for the rate you identified in question 1. Interpret what that confidence interval tells you._

###Part 2: Analyzing state food insecurity data###

We can also use the "STATE" variable to calculate rates for each state. To do this, we'll need to use the _spread_ function from tidyverse. Spread converts "long" data to "wide." Here's an example using the immigration data. First, let's filtering all records with a 0 and use ifelse to create a dummy variable for all individuals immigrating in the year 2000 or after.

```{r}
cps_data_2000<-cps_data %>%
  filter(YRIMMIG!=0) %>%
  mutate(yr_2000=ifelse(YRIMMIG>=2000,1,0)) 

cps_data_2000
```

Then we can use group_by, summarise, and spread to count the number of immigrants before and after 2000.

```{r}
cps_data_2000_wide<-cps_data_2000 %>%
  group_by(STATE,yr_2000) %>%
  summarise(count=n()) %>%
  spread(yr_2000,count)

cps_data_2000_wide
```

To convert this to a rate, we can use mutate with the above function. Here's the percentage of the immigrant population arriving since 2000.  Note that in this case, since your variables will be numbers, R requires tags around them: `1`, not just 1.

```{r}
cps_data_2000_wide<-cps_data_2000 %>%
  group_by(STATE,yr_2000) %>%
  summarise(count=n()) %>%
  spread(yr_2000,count) %>%
  mutate(total=`0` + `1`,
         rt2000=`1`/total*100)

cps_data_2000_wide
```

**Question 3 (5 points)** _Adapting the above function, create an estimated food insecurity rate for each state from these data. To do so, you'll need to create counts for each response (food secure, low food security, very low food insecurity), sum the latter two, and divide by the total responses within each state._

**Question 4 (3 points)** _Create a command that estimates the margin of error for each state based on the national rate you calculated in question 1 and the total responses for each state._

**Question 5 (3 points)** _Compare the margin of error you calculated for Georgia to the national margin of error. How do they differ? Mathematically, why are they different?_

**Question 6 (5 points)** _Create a column in your state food insecurity estimates that converts each state's food insecurity rate to a z score based on the whole population. What are the z scores for Wisconsin, Washington, and Mississippi? What do those z scores tell you?_

###Part 3: Sampling###

A new study is being developed to determine whether new food shelves in the Atlanta metropolitan area are reducing rates of food insecurity. The research question is whether living within a mile of a food pantry lowers food insecurity for households.

**Question 7 (4 points)** _Pick a probabilistic sampling strategy (or combination of strategies) discussed in our text or in lecture that would be appropriate for this research question: random, systemic, stratified, and cluster. Describe how this strategy could be used to create a sample for use in this proposed study. Describe one strength and weakness of this approach._

**Question 8 (3 points)** _Health officials would like to do a related survey of household food insecurity with enough responses to allow for margins of error under 2% (with 95% confidence). Assume that the rate is similar to the one you identified for Georgia in question 5.  Use R to compute how big a sample they would need._ 



