---
title: "Moran's I and Getis-Ord in R"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(sf)
library(albersusa)
library(tmap)
```

## Creating spatial weights

Moran's I is a measure of global spatial autocorrelation. The spdep package in R includes tools for this as well as local measures such as Getis-Ord Gi* and LISA. 

To start with, we'll create spatial weights. These define the neighbors of each county. The function below defines neighbors using "queen" contiguity, sharing an edge or corner. Non-contiguous polygons can be troublesome, so for this example, we'll just look at Georgia and its neighbors.

```{r}
#install.packages("spdep")
library(spdep)

election<-read_csv("https://github.com/jshannon75/geog4300/raw/master/data/elections0816_demog_pct.csv") %>%
  filter(year=="2016")

election_sf<-counties_sf() %>%
  filter(state %in% c("Georgia","Alabama","Tennessee","North Carolina",
                      "South Carolina","Florida")) %>%
  st_transform(32615) %>%
  rename(fips_code=fips) %>%
  inner_join(election)

election_nb<-poly2nb(election_sf,queen=TRUE)
```

This neighbors object is a list showing the row numbers of neighbors for each polygon. For example, to see the neighbors of the county in the first row (Abbeville County, AL), you'd just call it. The results show that the counties in rows 4, 7, 114, 121, and 127 are the neighbors. We can subset the full dataset by row number to see which counties those are.

```{r}
election_nb[[1]]

neighbors<-election_sf[election_nb[[1]],]
```

Once you have that list of neighbors you need to weight them, meaning determine what share of the average neighbor's score they receive. Often each neighbor is just weighted equally. But there are circumstances where each would have its own weight. For example, inverse distance weighting means close neighbors are weighted more heavily than far ones. Look at the help file for nb2listw for more information.

```{r}
election_weight<-nb2listw(election_nb,style="W",zero.policy=TRUE)
```

## Running the Moran's I test

Once weights are created, it's simple to run a Moran's I test. Are GOP votes spatially clustered?

```{r}
moran.test(election_sf$gop_pct,election_weight)
```

The Moran's I statistic comes back at 0.48, which is reasonably high. Like correlation, this statistic runs from 0 to 1. There's a p value calculated for this number, but a more robust measure is a Monte Carlo simulation--basically shuffling values around many times and seeing how common the observed outcome is. That looks like this:

```{r}
moran.mc(election_sf$gop_pct,election_weight,nsim=499)
```

##Getis-Ord Gi*

There's several measures of local spatial autocorrelation, which is often called hot spot analysis. While Moran's I gives a measure of overall clustering, these measures identify local high or low values. Getis-Ord Gi* identifies just these high and low clusters. 

Since we have a weights file already set up, we just need to call the function. The resulting numbers give a Z-score for each county. Values above 1.96 are significant hot spots (p=0.05), and values lower than -1.96 are cold spots.
 
```{r}
local_g<-localG(election_sf$gop_pct,election_weight)
election_sf$localG<-local_g

tm_shape(election_sf)+
  tm_polygons("localG")
```

If we wanted to just see hot or cold spots, we could recode the variable using case_when from the tidyverse.

```{r}
election_sf<-election_sf %>%
  mutate(localG_recode=case_when(localG <= -1.96~"Cold",
                                 localG < 1.96~"No value",
                                 localG >= 1.96~"Hot"))

tm_shape(election_sf)+
  tm_polygons("localG_recode")
```

