---
title: "Geog6300: Chi square tests"
output: html_notebook
editor_options: 
  chunk_output_type: console
---

Chi-square tests. First, we create the data and then run the test. Here we use the Thanksgiving.chisquare dataset from lecture.

```{r}
white<-c(15, 25)
dark<-c(25, 10)
tofurkey<-c(3, 7)
thanksgiving<-data.frame(cbind(white, dark, tofurkey))

chisq.test(thanksgiving)
```

Here's the same procedure for our in-class problem.

```{r}
car<-c(58, 199, 205)
transit<-c(21, 49, 32)
walk<-c(36, 64, 29)
survey<-data.frame(cbind(car, transit, walk))
chisq.test(survey)
```

Creating a contingency table is possible with tidyverse tools. Is there a difference in the severity of tornados (MAG) across states (ST)?  

```{r}
library(tidyverse)
tornado<-read_csv("data/tornado_pointsSE.csv")

tornado_table<-tornado %>%
  count(ST,MAG,name = "count") %>%
  pivot_wider(names_from=MAG,values_from=count,values_fill=0) %>%
  select(-`-9`) 
```

We can then do a chi-square on just the count variables, removing the state variable.

```{r}
chisq.test(tornado_table %>% select(-ST))
```

For interpretation, it can be better to transform the counts in this table to percentages. We can do so using pivot_wider, group_by, and mutate.

```{r}
tornado_table_pct<-tornado_table %>%
  pivot_longer(`0`:`5`,names_to="mag",values_to="count") %>% #Make longer
  group_by(ST) %>% mutate(total=sum(count)) %>% #Calculate total tornado count
  mutate(pct=round(count/total*100,1)) %>% #Calculate the percentage, rounded to 1 digit
  select(-count) %>% #Remove tornado count
  pivot_wider(names_from=mag,values_from=pct) #Make wider with percentages
```

Which states have the highest percentage of category 4 and 5 tornadoes? 
How would you describe the observed differences between states?

