---
title: "ANOVA and post-hoc tests"
output: html_notebook
editor_options: 
  chunk_output_type: console
---
```{r setup}
library(tidyverse)
library(broom)
```


##Post-hoc ANOVA analysis

Let's start by loading the "data" used in the video lecture.
```{r}
anova_data_navarro<-data.frame(part_id=c(4,5,6,13,14,15,8,9,16,17,18,7,1,2,3,10,11,12),
                               drug=c("anxifree","anxifree","anxifree","anxifree","anxifree","anxifree","joyzepam","joyzepam","joyzepam",
                                      "joyzepam","joyzepam","joyzepam","placebo","placebo","placebo","placebo","placebo","placebo"),
                               therapy=c("no.therapy","no.therapy","no.therapy","CBT","CBT","CBT","no.therapy","no.therapy",
                                         "CBT","CBT","CBT","no.therapy","no.therapy","no.therapy","no.therapy","CBT","CBT","CBT"),
                               result=c(0.6,0.4,0.2,1.1,0.8,1.2,1.7,1.3,1.8,1.3,1.4,1.4,0.5,0.3,0.1,0.6,0.9,0.3))
```

Next we create an ANOVA object and use summary to look at our results. What's our null hypothesis mean? How do we interpret a p<0.05? What other information is provided in the output?
```{r}
drugtest<-aov(result~drug,anova_data_navarro)
summary(drugtest)
```

We can also use the tidy function on ANOVA objects:

```{r}
drugtest_sf<-tidy(drugtest)
```

For post-hoc testing, we can use pairwise t test without and with Bonferroni correction.
```{r}
pairwise.t.test(x=anova_data_navarro$result, 
                g=anova_data_navarro$drug,
                var.equal=TRUE,
                pool.sd=FALSE) 

pairwise.t.test(x=anova_data_navarro$result, 
                g=anova_data_navarro$drug,
                var.equal=TRUE,
                pool.sd=FALSE,
                p.adjust.method="bonferroni")
```


Tukey HSD test is a more sophisticated test with correction. We use the ANOVA object for it. How is this output similar or different to pairwise.t.test?
```{r}
TukeyHSD(drugtest)
```



Testing residuals for normality--just use the "residuals" function on the ANOVA object. We can then create a qqplot and use shapiro.test to look at normality.
```{r}
drugtest.resid<-residuals(drugtest)
qqnorm(drugtest.resid);qqline(drugtest.resid) #Using a base R version of a QQ plot
hist(drugtest.resid)
shapiro.test(drugtest.resid)
```

Running a Fligner-Killeen test checks for equal variances between groups.
```{r}
fligner.test(result~drug,data=anova_data_navarro)
```

You could also treat the drug trial data as non-parametric. kruskal.test is the non-parametric version.
```{r}
kruskal.test(anova_data_navarro$result~anova_data_navarro$drug)
```

You try it! Test for differences in pollen levels for species across regions. Do any species NOT have regional differences? The region variable is Region_fac.
```{r}
pollen_data<-read_csv("data/Midwest_Pollen_Data.csv")
```

Some examples below...





















```{r}
pollen_aov<-aov(Picea~Region_fac,data=pollen_data)
summary(pollen_aov)
```

#Bonus: Functional programming
We could create a function that runs this analysis and then converts results to a tidy dataset using broom

```{r}
pollen_long<-pollen_data %>%
    pivot_longer(Picea:Herbsum,names_to="Pollen",values_to="Count")
region_aov<-function(pollen_type){
  pollen_sel<-pollen_long %>%
    filter(Pollen==pollen_type)
  
  aov_result<-aov(Count~Region_fac,data=pollen_sel)
  broom::tidy(aov_result) %>% 
    filter(term=="Region_fac") %>%
    mutate(type=pollen_type)
}

picea_aov<-region_aov("Picea")

results_all<-map_df(unique(pollen_long$Pollen),region_aov)
```





