---
title: "Kernel density and quadrats with spatstat"
output: html_notebook
editor_options: 
  chunk_output_type: console
---

For this script, you'll need tidyverse and sf, as well as a new (to us) package called spatstat. We're also going to use a couple of other spatial packages--`rgdal` and `maptools`. `rgdal` is an older R package that ports the GDAL library (used in a LOT of GIS software) into R.

```{r}
#library(sf)
library(knitr)
#install.packages(c("spatstat","maptools"))
library(tidyverse)
library(spatstat)
library(sf)
library(maptools)
```

We will work with tornado data for this analysis. We will also load a spatial file (in geopackage format) for states in the Southeast. Spatstat requires *projected data*, not just raw WGS84 coordinates. This dataset thus gives coordinates in that projection format. 

The raw data are available at https://www.spc.noaa.gov/gis/svrgis/.

The `as.owin` package creats an operating window for spatstat to use. `as_Spatial` converts the sf data into the older SpatialPolygons format used by rgdal.

```{r}
tornado_points<-read_csv("data/tornado_pointsSE.csv") %>%
  st_as_sf(coords=c("x_utm","y_utm"),remove=FALSE,crs=32615) #Coordinate system is UTM 15

states<-st_read("data/states_se.gpkg") %>%
  st_transform(32615) #Transform to UTM 15
states1<-as.owin(as_Spatial(states))
```

Quadrat analysis in R often uses a package called spatstat. A lot of info on spatstat is available here: https://research.csiro.au/software/wp-content/uploads/sites/6/2015/02/Rspatialcourse_CMIS_PDF-Standard.pdf. With spatstat, you can create square quadrants of varying sizes

To use the spatstat package, you must first determine the maximum and minimum values for your lat/long coordinates (using the point where each tornado started). Here, we use the min and max functions to do so.

```{r}
maxlat<-max(tornado_points$y_utm)
minlat<-min(tornado_points$y_utm)
maxlon<-max(tornado_points$x_utm)
minlon<-min(tornado_points$x_utm)
```

Spatstat needs a "ppp" object, which is a point pattern dataset. To create that, use the ppp function, providing X and Y points plus the range of your X and Y coordinates.

```{r}
tornado_ppp<-ppp(tornado_points$x_utm,tornado_points$y_utm,
                 c(minlon,maxlon),c(minlat,maxlat))
Window(tornado_ppp)<-states1
plot(tornado_ppp)
```

If desired, we can add a *convex hull* polygon, which is the simplest polygon including all observed points. To do so, we create the initial plot, create that polygon with the `convexhull` polygon, and add it using the `lines` function.

```{r}
plot(tornado_ppp)

hull<-convexhull(tornado_ppp)
lines(hull$bdry[[1]]$x,hull$bdry[[1]]$y,lty=2)
```

We can use the density function to create a kernel density map.

```{r}
kernel_map<-density(tornado_ppp)
plot(kernel_map)
```

We can change the size of the "kernel" used to interpolate this map. Again, remember that the unit here is meters. Which is best?

We also add state boundaries to each map by using `add=TRUE`.

```{r}
plot(density(tornado_ppp,1000))
plot(states1,add=TRUE,border="white")

plot(density(tornado_ppp,5000))
plot(states1,add=TRUE,border="white")

plot(density(tornado_ppp,10000))
plot(states1,add=TRUE,border="white")

plot(density(tornado_ppp,50000))
plot(states1,add=TRUE,border="white")

plot(density(tornado_ppp,100000))
plot(states1,add=TRUE,border="white")

plot(density(tornado_ppp,500000))
plot(states1,add=TRUE,border="white")

```

Spatstat can also create a contour map. The lines indicate the value of the kernel density map.

```{r}
map<-contour(density(tornado_ppp,100000))
plot(states1,add=TRUE,lty=2)

```

Spatstat can create a quadrat map with square cells. The two numbers are the count of columns and rows.

```{r}
plot(quadratcount(tornado_ppp,10,5)) 
plot(quadratcount(tornado_ppp,15,10)) 
plot(quadratcount(tornado_ppp,6,3)) 
```

You can combine these maps with the add=TRUE parameter.
```{r}
kernel_map<-density(tornado_ppp,50000)
quad_map<-quadratcount(tornado_ppp,6,3)

plot(kernel_map,add=TRUE)
plot(quad_map, add=TRUE,col="grey")
```

### You try it!
Create a kernel density map of states in the Southeast showing the frequency of tornados rated F3 or higher. The Fujita rating is listed under the "MAG" variable. Add state boundaries.

```{r}

```

