---
title: "Probability distributions"
output: github_notebook
editor_options: 
  chunk_output_type: console
---

R has several built in functions for analyzing and simulating different probability distributions. For binomial distribution there are two options:

1) dbinom povides the probability of an exact outcome. For example, what is the probability of getting *exactly* one two pair hand of cards over five deals? We know that the probability of two pair is .048. (see http://www.math.hawaii.edu/~ramsey/Probability/PokerHands.html)

```{r}
dbinom(1,size=5,prob=.048)
```

What is the probability of NO two pair hands?

```{r}
dbinom(0,size=5,prob=.048)
```

2) pbinom provides the *cumulative probability* of an outcome--from no occurance to a given maximum.
What is the probability of getting one or LESS two pair hand of cards of five deals?

```{r}
pbinom(1,size=5,prob=.048)
```

What is the probability of 2 or more two pair hands in five deals?

```{r}
1-pbinom(1,size=5,prob=.048)
```

The same options can be used for a Poisson distibution: dpois and ppois. If Georgia has 21 tornadoes every year, what is the probability of exactly 18 tornadoes in a year? For dpois, the two numbers are the outcome you're interested in (18 tornados) and the lambda (21).

```{r}
dpois(18,21)
```

What is the probability of 18 or less tornadoes?

```{r}
ppois(18,21)
```

What is the probability of more than 18 tornadoes?

```{r}
1-ppois(18,21)
```

You can visualize these distributions using the "visualize" package

```{r}
#install.packages("visualize")
library(visualize)
```

Probability of 1 or more two pair hands in 5 draws.
```{r}
visualize.binom(stat=1,size=5,prob=.048,section="upper")  
```

Probability of 15 or less tornadoes
```{r}
visualize.pois(stat=15,lambda=21,section="lower")
```

Probability of more than 16 tornadoes
```{r}
visualize.pois(stat=17,lambda=21,section="upper")
```


You try it!
1) If the there are 12 rainy days in the average September in Athens, what's the chance of getting more than 15 rainy days this September?

2) For the past 100 years, a stream close to Athens has been measured at a gauging station. A USGS scientist has crunched these numbers and determined that the results are normally distributed, with a mean of 35.26 CFM and a standard deviation of 4.61 CFM. Calculate the probabilities the stream flow in a given year will exceed 42 and 48 CFM. (Note that dnorm and pnorm work just like pbinom and pbinom for normal distributions. Use the help function to learn more about how to use the mean and sd with this function.)

Answers below...


















Question 1:

```{r}
1-ppois(15,12) #0.156

#Using visualize
visualize.pois(stat=16,lambda=12,section="upper")
```

Question 2:
```{r}
#42 inches
1-pnorm(42,mean=35.26,sd=4.61)
visualize.norm(stat=42,mu=35.26,sd=4.61,section="upper")

#48 inches
1-pnorm(48,mean=35.26,sd=4.61)
visualize.norm(stat=48,mu=35.26,sd=4.61,section="upper")
```

