---
title: "Kernel density and quadrats"
output: github_notebook
editor_options: 
  chunk_output_type: console
---

For this script, you'll need tidyverse and sf, as well as a new (to us) package called spatstat. We're also going to use a couple of other spatial packages--rgdal and maptools.

```{r}
#library(sf)
library(knitr)
opts_knit$set(root.dir = "~/path/to/folder/")

#install.packages(c("spatstat","rgdal","maptools"))
library(tidyverse)
library(spatstat)
library(sf)
library(maptools)
```

We will continue work with tornado data for this analysis. We will also load a spatial file (in geopackage format) for states in the Southeast. Spatstat requires a projection, so we will use UTM 15.

```{r}
tornado_points<-read_csv("data/tornado_pointsSE.csv") %>%
  st_as_sf(coords=c("x_utm","y_utm"),remove=FALSE,crs=32615) #Coordinate system is UTM 15

states<-st_read("data/states_se.gpkg") %>%
  st_transform(32615) #Transform to UTM 15
states1<-as.owin(as_Spatial(states))
```

Quadrat analysis in R often uses a package called spatstat. A lot of info on spatstat is available here: https://research.csiro.au/software/wp-content/uploads/sites/6/2015/02/Rspatialcourse_CMIS_PDF-Standard.pdf. With spatstat, you can create square quadrants of varying sizes

To use the spatstat package, you must first determine the maximum and minimum values for your lat/long coordinates (using the point where each tornado started). Here, we use the min and max functions to do so.

```{r}
maxlat<-max(tornado_points$y_utm)
minlat<-min(tornado_points$y_utm)
maxlon<-max(tornado_points$x_utm)
minlon<-min(tornado_points$x_utm)
```

Spatstat needs a "ppp" object. To create that, use the ppp function, providing X and Y points plus the range of your X and Y coordinates.

```{r}
tornado_ppp<-ppp(tornado_points$x_utm,tornado_points$y_utm,
                 c(minlon,maxlon),c(minlat,maxlat))
Window(tornado_ppp)<-states1
plot(tornado_ppp)
```

We can use the density function to create a kernel density map.

```{r}
plot(density(tornado_ppp))

kernel_map<-density(tornado_ppp)
plot(kernel_map)
```

We can change the size of the "kernel" used to interpolate this map. Again, remember that the unit here is meters. Which is best?

```{r}
plot(density(tornado_ppp,1000))
plot(density(tornado_ppp,5000))
plot(density(tornado_ppp,10000))
plot(density(tornado_ppp,50000))
plot(density(tornado_ppp,100000))
plot(density(tornado_ppp,500000))
```

Spatstat can also create a contour map. The lines indicate the value of the kernel density map.

```{r}
map<-contour(density(tornado_ppp,100000),)
```

Spatstat can create a quadrat map with square cells. The two numbers are the count of columns and rows.

```{r}
plot(quadratcount(tornado_ppp,10,5)) 
plot(quadratcount(tornado_ppp,15,10)) 
plot(quadratcount(tornado_ppp,6,3)) 
```

You can combine these maps with the add=TRUE parameter.
```{r}
kernel_map<-density(tornado_ppp,50000)
quad_map<-quadratcount(tornado_ppp,6,3)

plot(kernel_map,add=TRUE)
plot(quad_map, add=TRUE,col="grey")
```