---
title: "Tidy text and text mining"
output: html_notebook
---

In this script, we'll be working with the tidytext package to get introduced to text mining. This closely follows the Text Mining with R book by Julia Silge and David Robinson, available online at https://www.tidytextmining.com/index.html

We're going to need two packages: gutenbergr, which provides access to free texts at the Gutenberg project, and tidytext.

```{r}
#install.packages("gutenbergr")
library(gutenbergr)

#install.packages("tidytext")
library(tidytext)

library(tidyverse)
```

Let's start by downloading Marx's Capital (id 46423 on the Gutenberg project) and breaking it down by word using the unnest_tokens function.


```{r}
capital<-gutenberg_download(46423) 

capital_words<- capital %>%
  unnest_tokens(word,text)

head(capital_words)
```

We can remove common "stop words" and then use group_by and summarise to count the most common ones.

```{r}
capital_wordcount<-capital_words %>%
  anti_join(stop_words) %>%
  group_by(word) %>%
  summarise(marx_count=n())
View(capital_wordcount)
```

What's the distribution of word frequency?

```{r}
ggplot(capital_wordcount,aes(x=marx_count)) + geom_histogram()

```

We can use geom_col to visualize these with ggplot. Arrange orders the data before we plot it.

```{r}
capital_graph<-capital_wordcount %>%
  top_n(20) %>% #Selects the 20 highest count words
  mutate(word=reorder(word,marx_count)) 

ggplot(capital_graph,aes(word,marx_count)) +
  geom_col() +
  coord_flip()
```

You try it! 
Download The Wealth of Nations (book 3300) from Gutenberg and identify the ten most common words in the book (excluding stop words).

Bonus challenge: Find words that are in both Smith and Marx and create a scatterplot that shows their frequency in both. Hint: you'll need to use an inner_join.

In the Silge and Robinson book, there's much more detail about can identify common word groups and identify the relationships between them.
