---
title: "T and Z scores in R"
output: html_notebook
editor_options: 
  chunk_output_type: console
---

Let's start with our random distribution of ACT scores. The following code simulates random population of 100 taking the ACT in 2011. The rnorm comes from base R, and it just generates a random distribution from a population with the listed characteristics.

```{r}
ACT.scores<-data.frame(rnorm(100, mean=21, sd=5.3))
names(ACT.scores)<-"scores"

mean(ACT.scores$scores)
sd(ACT.scores$scores)
```

To calculate z scores, simply put the formula for z scores into R (value - mean)/sd. 

What's the z score for an ACT score of 29? The mean is 21 and sd is 5.3.

```{r}
z.score<-(29-21)/5.3
z.score
```

You could even convert the whole dataset to z scores. The scale function does something simlar--creating z scores based on the mean and sd of the *sample.*

```{r}
library(tidyverse)
ACT.scores<-ACT.scores %>%
  mutate(zscore=(scores-21)/5.3)

ACT.scores$zscore_scale<-scale(ACT.scores)

hist(ACT.scores$scores)
hist(ACT.scores$zscore)
hist(ACT.scores$zscore_scale)
```

##These two provide different numbers. Why??

What's the ACT score associated with the 67th percentile? The qnorm function looks that up. It translates a probability to a data value.

```{r}
score<-qnorm(0.67,mean=21,sd=5.3)
score
```

T scores follow a similar logic. The function below creates a dataset of 10 variables with the t-distribution when df=10. The mean of this dataset is 0 and the standard deviation is 1.

```{r}
t.sample<-data.frame(rt(100, df=10))
names(t.sample)<-"tscore"
hist(t.sample$tscore)
```

You try it! If the average household in Athens is 3.6 miles away from the nearest Waffle House, and the sd is 1.2 miles, what distance defines the 10% of households furthest from a Waffle House? This assumes the data are normally distributed (though there's nothing normal about Waffle House...)

```{r}

```


Answer below...





















Answer:
```{r}
dist<-qnorm(0.9,mean=3.6,sd=1.2)
dist
```

