---
title: "T and Z scores in R"
output: html_notebook
editor_options: 
  chunk_output_type: console
---

```{r}
library(tidyverse)
```

Let's start with our random distribution of ACT scores. The following code simulates random population of 100 taking the ACT in 2011.

```{r}
ACT.scores<-data.frame(rnorm(100, mean=21, sd=5.3))
names(ACT.scores)<-"scores"
```

To calculate z scores, simply put the formula for z scores into R (value - mean)/sd. 

What's the z score for an ACT score of 29? The mean is 21 and sd is 5.3.

```{r}
z.score<-(29-21)/5.3
z.score
```

You could even convert the whole dataset to z scores. The scale function does something simlar--creating z scores based on the mean and sd of the sample.

```{r}
ACT.scores<-ACT.scores %>%
  mutate(zscore=(scores-21)/5.3)

ACT.scores$zscore_scale<-scale(ACT.scores)

hist(ACT.scores$scores)
hist(ACT.scores$zscore)
hist(ACT.scores$zscore_scale)
```

##These two provide different numbers. Why??

What's the ACT score associated with the 67th percentile? The qnorm function looks that up.

```{r}
score<-qnorm(0.67,mean=21,sd=5.3)
score
```

T scores follow a similar logic. The function below creates a dataset of 10 variables with the t-distribution when df=10. The mean of this dataset is 0 and the standard deviation is 1.

```{r}
t.sample<-data.frame(rt(100, df=10))
names(t.sample)<-"tscore"
hist(t.sample$tscore)
```

You try it! If the average household in Athens is 3.6 miles away from the nearest Waffle House, and the sd is 1.2 miles, what distance defines the 10% of households furthest from a Waffle House? This assumes the data are normally distributed (though there's nothing normal about Waffle House...)
