---
title: "Geog6300: Correlation"
output: html_notebook
editor_options: 
  chunk_output_type: console
---

```{r setup}
library(tidyverse)
#install.packages("Hmisc","broom","ggcorplot","corrgram")
library(Hmisc)
library(ggcorrplot)
library(corrgram)
```


Correlation

Here's the fictitious dataset we used in lecture.
```{r}
MHI<-c(40000, 59000, 61000, 66000, 68000, 71000, 78000, 82000, 89000, 103000, 104000,113000, 129000, 141000,156000)
VCR<-c(45.2, 44.6, 39.2, 42.4, 37.4, 29.4, 38.1, 28.5, 21.4, 19.4, 25.1, 28.5, 22.5, 17.3, 13.5)

crimedata<-as_tibble(cbind(MHI,VCR))
crimedata
```


A simple scatterpot shows possible correlation.

```{r}
plot(crimedata$MHI, crimedata$VCR)

#Same plot in ggplot
ggplot(crimedata,aes(x=MHI,y=VCR))+geom_point()
```

Use the cor function to see the correlation.

```{r}
cor(crimedata$MHI,crimedata$VCR)
```


Correlations are often shown as a matrix. Here, we can use the pollen data for the Midwest. The rcorr function in the Hmisc package does this well.

```{r}
Midwest_Pollen_Data<-read_csv("data/Midwest_Pollen_Data.csv")
pollen<-Midwest_Pollen_Data %>%
  select(precpyr,Picea:Betula)#Take just the first 5 species

pollen_cor<-rcorr(as.matrix(pollen))#The data must be formatted as a matrix
pollen_cor
```

Note that this shows both the correlation coefficient and its p value.

We can also extract these correlations using tidy from the broom package. You can sort and filter to see which associations are the most significant.

```{r}
library(broom)
cor_values<-tidy(pollen_cor)
```

What if we wanted to see the association of precipitation and spruce trees (picea) across two regions? We could subset these data and then run the analysis. We'll subset results so we just have correlation with precpyr

```{r}
pollen_reg1<-Midwest_Pollen_Data %>%
  filter(Region==1) %>%
  select(precpyr,Picea:Betula)

pollen_reg2<-Midwest_Pollen_Data %>%
  filter(Region==2) %>%
  select(precpyr,Picea:Betula)

cor_reg1<-rcorr(as.matrix(pollen_reg1),type="spearman")
cor_reg2<-rcorr(as.matrix(pollen_reg2),type="spearman")
cor_all<-bind_rows(tidy(cor_reg1) %>% mutate(region="1"),
                   tidy(cor_reg2) %>% mutate(region="2")) %>%
  filter(column2=="precpyr")
```

Now we can visualize those results in a ladder plot, which shows lines between two regions.

```{r}
ggplot(cor_all,aes(x=region,y=estimate,group=column1,color=column1))+
  geom_line()+
  geom_point()
```

You can visualize associations using the pairs function, which is part of "base R" (comes installed). This creates scatterplots for every variable pair.
```{r}
cor_all<-Midwest_Pollen_Data %>% 
        select(-Region,-Region_fac,-seqnum:-tmeanjul)

pairs(cor_all)
```

That's a lot! Let's just visualize those variables with correlations of 0.6 or more (or -0.6 or less).

```{r}

cor_values<-tidy(rcorr(as.matrix(cor_all)))

pollen_select<-cor_values %>%
  filter(estimate > 0.6 | estimate < -0.6) 
pollen_list<-unique(c(pollen_select$column1,pollen_select$column2))

pollen_filter<-Midwest_Pollen_Data %>%
  select(c(precpyr,pollen_list))

pairs(pollen_filter)
```

There's lots of ways to visualize correlation matrices, and we often want to look at lots of variable combinations at once. Here's ggcorrplot. (More info: http://www.sthda.com/english/wiki/ggcorrplot-visualization-of-a-correlation-matrix-using-ggplot2)

```{r}
correlation_matrix <- cor(pollen_filter)
ggcorrplot(correlation_matrix, 
           type = "upper", 
           method="circle")

ggcorrplot(correlation_matrix,
    lab=TRUE, #Include numbers
    type = "lower", 
    insig = "blank") #Leave out insignificant cells
```

The corrgram function in the corrgram package is another option. https://rawgit.com/kwstat/corrgram/master/vignettes/corrgram_examples.html

```{r}
corrgram(correlation_matrix, order=FALSE,
         lower.panel=panel.conf,
         upper.panel=panel.shade, text.panel=panel.txt,
         main="Census correlations")
```


For non-parametric data, specify spearman's rho in the rcorr package.
```{r}
rcorr(as.matrix(pollen), type="spearman")
```

You can also specify test for the cor command.
```{r}
M<-cor(pollen,method="spearman")
corrgram(M, panel=panel.shade)
```



