---
title: "Lecture 3-1: group_by, summarise, and ggplot"
output: html_notebook
---

The ggplot2 package allows you to do more elegant looking graphs. See this cheat sheet: https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf. It's a standalone package, but also part of the tidyverse

Let's load the census dataset we've been using.

```{r, message=FALSE}
library(tidyverse)
census_data<-read_csv("https://github.com/jshannon75/geog4300/raw/master/data/ACSCtyData_2014ACS.csv")
```

If we want to summarise data, we can just use the summarise function in the tidyverse. Let's look at the BADeg_Pct variable

```{r}
census_data_summary<-census_data %>%
  summarise(mean=mean(BADeg_pct),
            median=median(BADeg_pct),
            sd=sd(BADeg_pct),
            iqr=IQR(BADeg_pct),
            cv=sd/mean)
census_data_summary
```

The tidyverse also makes it easier to summarise by sub-groups. We just use the group_by function to identify the categorical variable(s) that define our groups.

```{r}
census_data_summary_grp<-census_data %>%
  group_by(Region) %>%
  summarise(mean=mean(BADeg_pct),
            median=median(BADeg_pct),
            sd=sd(BADeg_pct),
            iqr=IQR(BADeg_pct),
            cv=sd/mean)
census_data_summary_grp
```

We can group by multiple variables--here using both region and state names.

```{r}
census_data_summary_grp<-census_data %>%
  group_by(Region, St_name) %>%
  summarise(mean=mean(BADeg_pct),
            median=median(BADeg_pct),
            sd=sd(BADeg_pct),
            iqr=IQR(BADeg_pct),
            cv=sd/mean)
census_data_summary_grp
```

**You try it!**
Work on question 2 for lab 1!

###Using ggplot

ggplot is structured in a particular way:
* "ggplot(data,aes(x=variable...))" 

Different "geoms" specify you want to plot those data. See this list of available geoms: http://sape.inf.usi.ch/quick-reference/ggplot2/geom. While ggplot is part of the tidyverse, it uses the plus sign, not the pipe, to separate elements.

In this example, the second function, "geom_point," says you want a point scatterplot.

```{r}
ggplot(census_data, aes(x=afam_pop_pct,y=BADeg_pct))+
  geom_point()
```

In this example, the function is a histogram. Only one variable is needed. Remember that typing ? + the function will always bring up a help file.

```{r}
ggplot(census_data, aes(x=afam_pop_pct)) + 
  geom_histogram()
ggplot(census_data, aes(x=BADeg_pct)) + 
  geom_histogram()
```

Additional parameters can be added to the histogram and variable function to change the transparency (alpha) and "bin size" (binwidth) of the graph.

```{r}
ggplot(census_data, aes(x=BADeg_pct)) +
  geom_histogram(binwidth=.5, alpha=0.9)
ggplot(census_data, aes(x=BADeg_pct)) +
  geom_histogram(binwidth=4, alpha=0.3)
ggplot(census_data, aes(x=BADeg_pct)) +
  geom_histogram(binwidth=2, alpha=0.5)
```

Density graphs smooth histograms. In this case, the fill parameter below specifies what the fill color should be, using a hex code for the color, and makes the graph semi-transparent using the alpha parameter.

```{r}
ggplot(census_data, aes(x=BADeg_pct)) + 
  geom_density(fill="#0072B2",alpha=0.5)
```

ColorBrewer color schemes can be used with variables. Load the RColorBrewer package. See also the ColorBrewer site: http://colorbrewer2.org/. See the RColorBrewer documentation for palette names, or look at this guide: https://www.nceas.ucsb.edu/~frazier/RSpatialGuides/colorPaletteCheatsheet.pdf.

In the example below, we use "fill" with ggplot to define a categorical variable that breaks the data into subgroups, which can be visualized using the specifed color scheme in the scale_fill_brewer function.

```{r}
ggplot(census_data, aes(x=BADeg_pct, fill=Region)) + 
  geom_histogram() + 
  scale_fill_brewer(palette="Set1")
ggplot(census_data, aes(x=BADeg_pct, fill=Region)) + 
  geom_histogram() + 
  scale_fill_brewer(palette="RdGy")

```

**You try it!**
Make a histogram for pov_pop_pct (% in poverty) changing colors by region. Use the Dark2 ColorBrewer palette.


You can create boxplots using geom_boxplot.

```{r}
ggplot(census_data, aes(y=BADeg_pct, x=1))+
  geom_boxplot()
```

Adding a second categorical y variable allows you to break a boxplot into parts.

```{r}
ggplot(census_data, aes(y=BADeg_pct, x=Region))+
  geom_boxplot()
```

**You try it**
Create the plot for question 5 in Lab 1!

We can also use facet wrap and facet grid to break these plots up. Here's a scatterplot with a smoothed trend line.

```{r}
ggplot(census_data, aes(x=afam_pop_pct,y=BADeg_pct))+
  geom_point()+
  geom_smooth()
```

Now let's break that into regions. The first graph has one chart with the points colored by region (see the parameter in geom_point). The second uses facet_wrap. The "scales='free_x'" parameter allows the x axis to vary in each graph. 

```{r}
ggplot(census_data, aes(x=afam_pop_pct,y=BADeg_pct))+
  geom_point(aes(color=Region))+
  geom_smooth()

ggplot(census_data, aes(x=afam_pop_pct,y=BADeg_pct))+
  geom_point()+
  geom_smooth()+
  facet_wrap(~Region, scales="free_x")
```

**You try it**
Work on the plots for questions 7 and 8 in lab 1.