---
title: "Geog4/6300: Clustering"
output: html_notebook
editor_options: 
  chunk_output_type: console
---

```{r setup}
library(tidyverse)
library(tmap)
library(sf)
```

##K-means Clustering
Let's read in our county level census data.

```{r}
countydata<-st_read("data/ACSCtyData_2014ACS_simplify.gpkg",stringsAsFactors=FALSE)
```

We can cluster these counties based on demographic similarity. Let's just look at educational attainment, for example. Here's how you would use kmeans to do so. First we remove all other variables except the ones of interest. We'll also have to remove the geometry column.

```{r}
county_ed<-countydata %>%
  select(LESSHS_PCT:GRADDEG_PC) %>%
  st_set_geometry(NULL)

ed_cluster<-kmeans(county_ed,5)
```

This produces an object with multiple components. We just want the cluster number, so we'll extract that, join it to the data frame, and then add the geometry column by joining it back to the initial dataset. For ease of mapping, let's remove Alaska and Hawaii.

```{r}
clusters<-data.frame(ed_cluster$cluster)
names(clusters)="cluster"

countydata_cluster<-county_ed %>%
  bind_cols(clusters) 

countycluster_sf<-countydata %>%
  left_join(countydata_cluster) %>%
  mutate(cluster_char=as.character(cluster)) %>%
  filter(State!="Alaska" & State!="Hawaii")
```

Let's map out the clusters that result.

```{r}
tm_shape(countycluster_sf) + 
  tm_polygons("cluster_char")

tmap_mode("view")
tm_shape(countycluster_sf) + 
  tm_polygons("cluster_char",alpha=0.3)
```

What do these clusters mean? We could summarise them by looking at the mean values for each variable by cluster.

```{r}
cluster_vars<-countycluster_sf %>%
  st_set_geometry(NULL) %>% #Remove the geometry column
  select(cluster_char,LESSHS_PCT:GRADDEG_PC) %>% #Select variables if interest
  gather(LESSHS_PCT:GRADDEG_PC,key=var,value=value) %>% #Make data long for the summary
  group_by(cluster_char,var) %>%
  summarise(var_mean=mean(value)) 
cluster_vars
```

We can graph the results. If there's not a clear "theme" to each cluster, that's one sign the number of clusters isn't right.

```{r}
ggplot(cluster_vars,aes(x=var,y=var_mean,fill=cluster_char))+
  geom_bar(stat="identity",position="dodge")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

#Hierarchical clustering
For hierarchical clustering, we use the `hclust` function in base R. First, we need to calculate a distance matrix, showing the difference between observations. Then we use hclust to group observations at different levels based on difference. We can plot the resulting dendrogram. The hca_clust function draws boxes showing how different numbers of clusters (set using k=?) would be grouped.

```{r}
ed_dist<-dist(county_ed)

hca_clust<-hclust(ed_dist)
plot(hca_clust)
rect.hclust(hca_clust, k = 5)
```

The `cutree` function then creates a list of clusters for each observation.

```{r}
groups<-data.frame(cutree(hca_clust,k=5))
names(groups)<-"hca_clust"

countyed_hca<-bind_cols(county_ed,groups) 
countycluster_hca_sf<-countydata %>%
  left_join(countyed_hca)%>%
  mutate(hcaclust_char=as.character(hca_clust)) %>%
  filter(State!="Alaska" & State!="Hawaii")

tm_shape(countycluster_hca_sf)+
  tm_polygons("hcaclust_hcar")
```

What do these clusters look like?

```{r}
cluster_vars_hca<-countycluster_hca_sf %>%
  st_set_geometry(NULL) %>% #Remove the geometry column
  select(hcaclust_char,LESSHS_PCT:GRADDEG_PC) %>% #Select variables if interest
  gather(LESSHS_PCT:GRADDEG_PC,key=var,value=value) %>% #Make data long for the summary
  group_by(hcaclust_char,var) %>%
  summarise(var_mean=mean(value)) 
View(cluster_vars_hca)

ggplot(cluster_vars_hca,aes(x=var,y=var_mean,fill=hcaclust_char))+
  geom_bar(stat="identity",position="dodge")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

How many clusters is right? There's seldom a clear answer. See this page for a walkthrough of some computational methods that can be of use: https://www.datanovia.com/en/lessons/determining-the-optimal-number-of-clusters-3-must-know-methods/

You try it! Pick four variables of interest and use one of these methods to create clusters.