---
title: "Data manipulation: spread, gather, and join; Loading vector data"
output: html_notebook
editor_options: 
  chunk_output_type: console
---

In this script, we'll go over learn a few more functions from the tidyverse: gather, spread, left_join, and anti_join. We'll start by calling the tidyverse package and reading in the county census data we've worked with before.

```{r}
library(tidyverse)

census_data<-read_csv("https://github.com/jshannon75/geog4300/raw/master/data/ACSCtyData_2014ACS.csv")
```

###Group_by and summarise
What if we wanted to look at some summary statistics by state? The easiest way to do this in the tidyverse is using group_by and summarise. This allows us to calculate statistics for a numeric interval using some kind of qualitative grouping variable (state name, for example). Here's an example, summarizing the median value for BA degree attainment. In this case, we group by the state name variable and then use summarise to calculate the mean. 

Note that summarise is similar to, but different from mutate. It condenses multiple observations down into a single summary variable, rather than keeping them as separate observations.

```{r}
badeg_state<-census_data %>%
  group_by(St_name) %>%
  summarise(BA_mean=mean(BADeg_pct))

badeg_state2<-census_data %>%
  group_by(St_name) %>%
  mutate(BA_mean=mean(BADeg_pct)) %>%
  select(cty_name,St_name,BADeg_pct,BA_mean)
```


###Long and wide format
Summarise multiple variables can be tricky. You can calculate each mean separately, but this can be cumbersome for a large number of variables. One solution is to use functions that collapse multiple variables down to just two columns: one for the variable name and the other for the observational values. This "long" format is useful in lots of ways, but for our purposes, it can allow us to summarise multiple variables in just one line of code.

For awhile, the main commands that are used for reshaping datasets in this way have been 'gather' and 'spread.' The tidyverse developers have talked abotu replacing these with similar functions called 'pivot_longer' and 'pivot_wider', but those haven't been fully rolled out yet. So we'll stick with gather/spread 

Here's how they work. Let's say we wanted to summarise all the education variables. We could first select just the variables of interest and then use 'gather' to *gather* all these variables into two columns. We use the parameters key and value to define the column names for these columns. Here's what this looks like.

```{r}
edvars<-census_data %>%
  select(cty_name,St_name,LessHS_pct:GradDeg_pct) %>%
  gather(LessHS_pct:GradDeg_pct,key="var",value="values")
View(edvars)
```

You'll notice that this new table has just four columns, the county and state names, and then the variable and values for all of those education variables. We can now use group_by and summarise to calculate the mean values.

```{r}
edvars1<-edvars %>%
  group_by(St_name,var) %>%
  summarise(mean_value=mean(values))
```

Easy, right? But the resulting table is hard to read. We can now use 'spread' to *spread* out the data. In this case, we just have to identify the variable name and value columns.

```{r}
edvars2<-edvars1 %>%
  spread(var,mean_value)
```

If we prefer, we can do all this in one long, piped chain:

```{r}
edvars3<-census_data %>%
  select(cty_name,St_name,LessHS_pct:GradDeg_pct) %>%
  gather(LessHS_pct:GradDeg_pct,key="var",value="values") %>%
  group_by(St_name,var) %>%
  summarise(mean_value=mean(values)) %>%
  spread(var,mean_value)
```

Now you try it! Calculate the mean value by state for the nat_ins_pct, fbnat_ins_pct, and fb_ins_pct variables.

###Introducing joins

If you've used a GIS desktop package, you should be familiar with the idea of joins. Based on some common variable, joins provide a way to merge two datasets together. The tidyverse has LOTS of ways to join. Here's a couple of summaries of them all: 
* https://medium.com/@HollyEmblem/joining-data-with-dplyr-in-r-874698eb8898
* https://mikoontz.github.io/data-carpentry-week/lesson_joins.html

The most common, at least in my experience, is the left_join. When joining datasets A and B, this command keeps all rows in A and includes matching values from B. It will NOT add new values from B that don't have a match in A--you'd want right_join or full_join for that.

Here's an example. Let's say we want to add the poverty rate to our summary of educational attainment above. We can do so using group_by and summarise:

```{r}
povrate_st<-census_data %>%
  group_by(St_name) %>%
  summarise(pov_pop_pct=mean(pov_pop_pct))
```

To connect this new variable to our existing data frame, we can use left_join. As long as the matching variable (St_name in this case) is the same in both datasets, the function will detect it automatically. You'll see what variables it used in red text. It looks like this:

```{r}
edpov<-edvars3 %>%
  left_join(povrate_st)
```

Joins are really handy functions when combining datasets.

