---
title: "Data manipulation: group_by, summarise, and pivoting"
output: html_notebook
editor_options: 
  chunk_output_type: console
---

In this script, we'll go over learn a few more functions from the tidyverse: group_by, summarise, pivot_wider,pivot_longer, and left_join. We'll start by calling the tidyverse package and reading in the county census data we've worked with before.

```{r}
library(tidyverse)

census_data<-read_csv("data/ACSCtyData_2019ACS.csv")
```

###Group_by and summarise
What if we wanted to look at some summary statistics by state? The easiest way to do this in the tidyverse is using group_by and summarise. This allows us to calculate statistics for a numeric interval using some kind of qualitative grouping variable (state name, for example). Here's an example, summarizing the median value for BA degree attainment. In this case, we group by the state name variable and then use summarise to calculate the mean. 

Note that summarise is similar to, but different from mutate. It condenses multiple observations down into a single summary variable, rather than keeping them as separate observations.

```{r}
badeg_state<-census_data %>%
  group_by(St_name) %>%
  summarise(BA_mean=mean(BADeg_pct))

badeg_state2<-census_data %>%
  group_by(St_name) %>%
  mutate(BA_mean=mean(BADeg_pct)) %>%
  select(cty_name,St_name,BADeg_pct,BA_mean)
```

Now you try it! Calculate the median, mean, and standard deviation for the county poverty rate (pov_pop_pct) by state.

###Long and wide format
Summarizing multiple variables can be tricky. You can calculate each mean separately, but this can be cumbersome for a large number of variables. One solution is to use functions that collapse multiple variables down to just two columns: one for the variable name and the other for the observational values. This "long" format is useful in lots of ways, but for our purposes, it can allow us to summarize multiple variables in just one line of code. The two commands we use for this in the tidyverse are pivot_wide and pivot_longer.

Here's how they work. Let's say we wanted to summarize all the education variables. We could first select just the variables of interest and then use 'pivot_longer' to bring all these variables into two *longer* columns. The *names_to* parameter (on l. 43) names the variable with all the former variable headings. The *values_to* parameter (l. 44) tells what to do with all the values in those columns. Here's what this looks like.

```{r}
edvars<-census_data %>%
  select(cty_name,St_name,LessHS_pct:GradDeg_pct) %>%
  pivot_longer(LessHS_pct:GradDeg_pct,
               names_to="variable",
               values_to="rate")
View(edvars)
```

You'll notice that this new table has just four columns, the county and state names, and then the variable and values for all of those education variables. We can now use group_by and summarise to calculate the mean values.

```{r}
edvars1<-edvars %>%
  group_by(St_name,variable) %>%
  summarise(mean_rate=mean(rate))
```

Easy, right? But the resulting table is hard to read. We can now use 'pivot_wider' to push out the data out *wide*. In this case, we just have to identify the variable with the new variable names (using 'names_from') and values (using 'values_from').

```{r}
edvars2<-edvars1 %>%
  pivot_wider(names_from=variable,
              values_from=mean_rate)
```

If we prefer, we can do all this in one long, piped chain:

```{r}
edvars3<-census_data %>%
  select(cty_name,St_name,LessHS_pct:GradDeg_pct) %>%
  pivot_longer(LessHS_pct:GradDeg_pct,
         names_to="variable",
         values_to="rate") %>%
  group_by(St_name,variable) %>%
  summarise(mean_rate=mean(rate)) %>%
  pivot_wider(names_from=variable,
              values_from=mean_rate)
```

Now you try it! Calculate the mean value by state for the nat_ins_pct, fbnat_ins_pct, and fb_ins_pct variables.
