---
title: "Geog6300: Normality/non-parametric tests"
output: html_notebook
editor_options: 
  chunk_output_type: console
---

```{r setup}
library(tidyverse)
```


Normality tests and Wilcoxon non-parametric tests. This code chunk generates a skewed datsaet and then creates a qq plot of that dataset.

```{r}
test1<-rbeta(5000,7,2) 
hist(test1) 
qqnorm(test1) 
```

This code chunk does the same with a normal dataset.

```{r}
test2<-rnorm(5000,0,1)
hist(test2) 
qqnorm(test2) 
```

Shapiro-Wilk test: the null hypothesis is that the data ARE normal, so a low p value indicates that the dataset is skewed/not normal.
```{r}
shapiro.test(test1) #skewed data
shapiro.test(test2) #normal data
```

The qqline function can also add in a line showing normal distribution. It's helpful for seeing how much the actual data deviates.

```{r}
qqnorm(test1); qqline(test1)
qqnorm(test2); qqline(test2)
```

##You try it!
Go back to our census data (ACSCtyData_2014_ACS). 
```{r}
acsdata<-read_csv("data/ACSCtyData_2014ACS.csv")
```
What's the most normal variable you can find?


###Wilcoxon tests
The wilcox.test command does both Wilcoxon tests. Here's the rank sum example using lecture data. Think about what the null hypothesis is here. What does a significant p-value tell you?
```{r}
el.nino<-c(0, 5,1,3, 4, 1, 2, 2, 0, 2, 4)
la.nina<-c(4, 4,12, 2, 17, 7, 6, 10, 4, 8, 6)
wilcox.test(el.nino, la.nina)
```

The wilcox.test command results in a slightly different statistic than the formula used in the text/in class. It's actually computing an equivalent test called the Mann-Whitney test. The p-value can still be used in this case, and in both the finding is significant.

Wilcox.test can also compute *signed rank* tests.
```{r}
data2000<-c(3, 11, 9, 14, 17, 7, 21, 13, 19, 5)
data2010<-c(6, 12, 8, 18, 15, 13, 25, 9, 24, 10)
wilcox.test(data2000, data2010, paired=TRUE)
```
Here again, the Mann-Whitney test is used so the output is slightly different.

##You try it!
Let's look at a sample dataset of pollen reading for major tree species at several sites in the upper Midwest.

```{r}
pollen_data<-read_csv("data/Midwest_Pollen_Data.csv")
```
Pick two species and determine their normality. Compare their pollen levels using the appropriate t-test.
