---
title: "Geog4/6300: Multivariate Regression"
output: html_notebook
editor_options: 
  chunk_output_type: console
---

```{r setup}
#install.packages(c("car","lmtest"))
library(car)
library(tidyverse)
library(tmap)
library(lmtest)
library(sf)
library(Hmisc)
library(sjPlot)
library(corrplot)
```

##Multivariate regression
Multivariate regression uses the same command as univariate regression. Just add more variables. 

Let's start by loading the data illustrated in the class video: average date of the last spring frost:

```{r}
dlsf<-read_csv("data/DLSF_data.csv") %>%
  mutate(elev100=Elevation/100,
         dist50=DisttoCoas/50) %>%
  st_as_sf(coords=c("Longitude","Latitude"),crs=4326,remove=FALSE)

tmap_mode("view")
tm_shape(dlsf)+
  tm_bubbles(col="AvgDLSpF")
```

Now we can do univariate regression for each variable.

```{r}
model_lat<-lm(AvgDLSpF~Latitude,data=dlsf)
model_elev<-lm(AvgDLSpF~elev100,data=dlsf)
model_dist<-lm(AvgDLSpF~dist50,data=dlsf)

summary(model_lat)
summary(model_elev)
summary(model_dist)
```

To do a multivariate regression, simply add more dependent variables using `+`.
```{r}
model_all<-lm(AvgDLSpF~Latitude+elev100+dist50,data=dlsf)
summary(model_all)

tab_model(model_lat,model_elev,model_dist,model_all)
```

You can check for multicollinearity using the VIF test (variance inflation factor). The `vif` function is in the car package. You want values below 5
```{r}
vif(model_all)
```

You can also just look at the correlation matrix.
```{r}
dlsf_corr<-dlsf %>%
  select(AvgDLSpF,Latitude,Elevation,DisttoCoas) %>%
  st_set_geometry(NULL)
corrplot(cor(dlsf_corr))
```

We can check for heterosktasticity using the bptest function from the lmtest package. The null hypothesis is that the data are NOT heteroskedastic.
```{r}
bptest(model_all)
```

Let's test for outliers. We can plot the Cook's distance generally using the code below. Reminder: the numbers for the points refer to the row numbers of the observations in the data frame.

```{r}
cooks_dist<-cooks.distance(model_all)

cooks_dist<-data.frame(cooks=cooks.distance(model_all),
                       x=1:nrow(dlsf))
cooks_mean<-mean(cooks_dist$cooks)

ggplot(cooks_dist,aes(x=x,y=cooks,label=x))+
  geom_point()+
  geom_text(hjust=-0.2, vjust=0)+
  geom_hline(yintercept=cooks_mean*4,col="red")
```

Let's add a dummy variable for those outliers in the original dataset.

```{r}
cutoff<-mean(cooks_dist$cooks)*4

dlsf_outliers<-dlsf %>%
  bind_cols(cooks_dist) %>%
  mutate(outlier=if_else(cooks>cutoff,1,0)) 

tm_shape(dlsf_outliers) +
  tm_dots("outlier")
```

There are many ways to identify and deal with outliers. You can (1) top or bottom code them, (2) interpolate values from your model, (3) add additional control variables, or ($) remove observations if you feel they are spurious. 

One easy option is to see if the model results change much if you just remove the outliers. This allows you to assess whether they change the "headline story" of your model. If they don't, you may choose to leave them be. 

```{r}
dlsf_no_outlier<-dlsf_outliers %>%
  filter(outlier==0)

model_all1<-lm(AvgDLSpF~Latitude+elev100+dist50,data=dlsf_no_outlier)
summary(model_all1)
summary(model_all)

tab_model(model_all,model_all1)
```

If we add state names as an additional variable, R will automatically create dummy variables for each, keeping whatever one is first as a reference. We can switch the reference by changing this variable to a factor and listing the order.

```{r}
model_all_st<-lm(AvgDLSpF~Latitude+elev100+dist50+State,data=dlsf)
summary(model_all_st)

tab_model(model_all,model_all_st)

#Make Florida the reference category
dlsf1<-dlsf %>%
  mutate(State=factor(State,levels=c("FL","AL","GA","NC","SC","VA")))

model_all_st1<-lm(AvgDLSpF~Latitude+elev100+dist50+State,data=dlsf1)
summary(model_all_st1)

tab_model(model_all_st,model_all_st1)
```

