---
title: "Data manipulation: spread, gather, and join; Loading vector data"
output: html_notebook
---

In this script, we'll go over loading spatial data using the sf package, revisit tidycensus, and learn a few more functions from the tidyverse: gather, spread, left_join, and anti_join. Let's start by loading three packages. You've used tidyverse and tidycensus before, but sf is a main package for working with spatial data.

```{r}
library(tidyverse)
library(tidycensus)
library(sf)
```

The sf package uses the st_read function to read vector data in a variety of formats. You can use it to read in county boundaries for example:

```{r}
counties<-st_read("https://github.com/jshannon75/geog4300/raw/master/data/ACSCtyData_2014ACS_simplify.gpkg") %>%
  filter(State=="Georgia") %>%
  select(CTY_NAME,GEOID) 
```

One of the main advantages of sf is that it creates data frames that work well with the tidyverse functions. In the above case, we download all counties, select only those in Georgia, and then select just a county name and FIPS variable.

###Gather and spread

Let's work with the American Community Survey dataset. We are downloading two variables related to housing tenure: total households and renter occupied households.

```{r}
v15 <- load_variables(2016, "acs5", cache = TRUE)

census_vars<-c("B25007_001E","B25007_012E")

census_data<-get_acs(geography="county",variables=census_vars,state="GA") 
head(census_data)
```

This data comes in "long" format, meaning that the three variables are listed sequentially in separate rows. That's fine if we want to just filter for each individually, but not so good if we want to use these to calculate rates, such as percent renter occupied.

Let's filter out counties with high uncertainty. There's lots of ways we could define this, but for this walkthrough we'll use a basic one--counties where the margin of error for rental households is more than 10% of the estimate.

```{r}
census_data1<-census_data %>%
    filter(moe/estimate<0.1)
```

We can then use the spread function to make the data "wide," removing the margin of error column in the process.

```{r}
census_data_wide<-census_data1 %>%
  select(-moe) %>%
  spread(variable,estimate) 
head(census_data_wide)

```

There's lots of NAs in the resulting table, as many columns don't have reliable values for all three variables. So we can filter just for counties that have values in all three.

```{r}
census_data_wide<-census_data_wide %>%
  filter(B25007_001>0 & B25007_012>0)
head(census_data_wide)
```

Lastly, we can use mutate to create a rental housing rate.

```{r}
census_data_pct<-census_data_wide %>%
  mutate(rent_pct=B25007_012/B25007_001*100)
```

The mirror function of spread is gather, which takes multiple columns and condenses them to two: a "key" column for variable names/column headings and a "value" column for values. The function lets you name those two columns. Here's an example.

```{r}
census_data_long<- census_data_pct %>%
  gather(B25007_001:rent_pct,key="variable",value="value")
head(census_data_long)
```

Now you try it! Lab 2 asks you to download variables related to means of commute to work. Download those data, compute the percentage based on the normalizing variable, and spread the variables to wide format.


###Joining data

We can also join these data to our county shapefile. Joins combine datasets based on a common variable, usually some kind of id field. In this case, we're going to use the GEOID field, which has the Census FIPS code. There's lots of ways to do a join. Here's a link to all the options: http://stat545.com/bit001_dplyr-cheatsheet.html. But in this case, we're going to look at left_join, right_join, and anti_join.

Left_join is perhaps the most common join function. It matches all the join fields in the first dataset to the corresponding join fields in the second dataset. If there's records in the second dataset that don't match, they just get dropped. It will also autodetect the join fields--matching columns with identical names. Here's what that looks like in our case:

```{r}
census_data_pct_left<-left_join(counties,census_data_pct)
View(census_data_pct_left)
```

If you open up the new data frame, you'll see that many counties have NA for the census variables, since they didn't have a match. The right_join function may be a better choice here, as it will only keep counties from the first dataset with a match in the second one.

```{r}
census_data_pct_right<-right_join(counties,census_data_pct)
View(census_data_pct_right)
```

Note that in our case, it's important for counties to remain as the first dataset, as that's the only way that the spatial data will be preserved.

We'll talk more about how to map with R, but for now, plot gives us an idea of what these data look like.

```{r}
plot(census_data_pct_right)
```

Lastly, anti_join keeps only those counties in the first dataset that do NOT have a match in the second. In this case, it can help us identify which regions of the state lack reliable census data.

```{r}
census_data_pct_anti<-anti_join(counties,census_data_pct)
View(census_data_pct_anti)
```

```{r}
plot(census_data_pct_anti)
```

You try it!
Join the commute data you downloaded earlier to the county dataset. Filter for those counties where 5% or more of the population use public transit.