---
title: "Spatial data in R"
output:
  html_document:
    df_print: paged
editor_options: 
  chunk_output_type: console
---

### Loading and manipulating spatial data

There are several packages available to create and edit spatial data in R. This includes both raster and vector data. This script focuses on the latter. The relatively new sf (stands for simple features) package is one efficent way to load vector data. Other popular packages for spatial data are rgdal, sp, and raster.


```{r, message=FALSE}
library(tidyverse)
library(tidycensus)
library(sf)
```

First, let's load total population and county boundaries using the tidycensus package.

```{r}
ctypop<-get_acs(geography="county",variable="B01001_001",geometry=TRUE) %>%
  st_transform(4326) 
```

We can also read in the centroids of tornado paths in the Southeast over the last 50 years. The st_as_sf function takes coordinates in the csv and converts them to spatial data.

```{r}
tornado<-read_csv("https://github.com/jshannon75/geog4300/raw/master/data/tornado_points.csv")
tornado_sf<-st_as_sf(tornado,
                  coords=c(x="CLONG",y="CLAT"), #Indicate the coordinate columns
                  crs=4326, #Set the projections
                  remove=FALSE) #Keep the coordinate columns
```

The sf package also has a set of spatial functions that can be used on these data. For example, you could convert these polygons to centroids. We then use geom_sf in ggplot to visualize those points.

```{r}
ctypop_GA<-ctypop %>%
  mutate(st_fips=substr(GEOID,1,2)) %>% #Select just the first 2 digits of GEOID
  filter(st_fips==13) 
ggplot(ctypop_GA) + geom_sf()

ctypop_GA_points<-st_centroid(ctypop_GA)
ggplot(ctypop_GA_points) + geom_sf()
```

We can use st_join to join the county id to every tornado point based on its location. We join county data based on whether the point is within the county polygon.

```{r}
tornado_cty<-tornado_sf %>%
  st_join(ctypop,join=st_within)

ggplot(tornado_cty) + geom_sf(aes(color=GEOID),show.legend=FALSE)
```

We can then remove the spatial data and tally these points by county and as a whole. 

```{r}
tornado_cty_df<-tornado_cty %>%
  st_set_geometry(NULL) #Remove the geographic data

tornado_cty_count<-tornado_cty_df %>% 
  group_by(GEOID,MONTH) %>%
  summarise(count=n()) %>%
  spread(MONTH,count,fill=0) %>%
  mutate(total=Jan+Feb+Mar+Apr+May+June+July+Aug+Sept+Oct+Nov+Dec)

tornado_all<-tornado_cty_df %>%
  group_by(MONTH) %>%
  summarise(count=n()) %>%
  spread(MONTH,count,fill=0) %>%
  mutate(total=Jan+Feb+Mar+Apr+May+June+July+Aug+Sept+Oct+Nov+Dec)
```

Here's your task! We've covered how location quotient works in class. Create a formula that calculates the location quotient for tornadoes in March in each county.

### Tools for viewing/mapping spatial data
```{r}
library(mapview)
library(tmap)
```

You can use ggplot to view spatial data with the geom_sf package. For example, here's the tornado count data in Georgia for March. Scale_fill_viridis allows you to use the color scheme from Matlab (viridis)

```{r}
tornado_ga<- ctypop %>%
  left_join(tornado_cty_count) %>%
  filter(substr(GEOID,1,2)==13)

ggplot(tornado_ga) + 
  geom_sf(aes(fill=Mar)) + 
  scale_fill_viridis_c()
```

There's now multiple other packages as well. You can use the tmap package, which provides lots of mapping tools. Here's a basic plot:

```{r}
tm_shape(tornado_ga)+
  tm_polygons()
```

You can make a choropleth map by adding a variable. Here, the county FIPS code is also included with tm_text.

```{r}
tornado_ga_march<-tornado_ga %>%
  select(GEOID,Mar,Apr,May)

tm_shape(tornado_ga_march)+
  tm_polygons("Mar")+
  tm_text("GEOID",size=0.5)
```

Or you can add the tornados as points with the population data. This also adds a north arrow and scale bar and moves the legend outside.

```{r}
tornado_sf_ga<-tornado_sf %>% filter(ST=="GA")

tm_shape(ctypop_GA)+
  tm_polygons("estimate",style="jenks")+
tm_shape(tornado_sf_ga)+
  tm_dots(size=0.1) +
tm_compass()+
tm_scale_bar(position="left")+
tm_legend(legend.outside=TRUE)
```

You can use tmap for small multiple maps. Here, we gather all the months and and select four in spring. Then we use tm_facets to facet wrap them.

```{r}
tornado_ga_march_long<-tornado_ga %>%
  gather(Apr:Sept,key="month",value="count") %>%
  filter(month %in% c("Mar","Apr","May","June"))

tm_shape(tornado_ga_march_long)+
  tm_polygons("count",style="jenks")+
  tm_facets(by="month")
```


You can also do this with ggplot. We use "factor" to order the months for our output.
```{r}
tornado_ga_long<-tornado_ga %>%
  gather(Apr:Sept,key="month",value="count") %>%
  mutate(month_f=factor(month,levels=c("Jan","Feb","Mar","Apr","May","June",
                                    "July","Aug","Sept","Oct","Nov","Dec")))

ggplot(tornado_ga_long) +
  geom_sf(aes(fill=count))+
  scale_fill_viridis_c()+
  facet_wrap(~month_f)
```


You can also make interactive maps with tmap.

```{r}
tmap_mode("view") #To shift back to static maps, use tmap_mode("plot")

tm_shape(ctypop_GA)+
  tm_polygons("estimate",style="jenks")+
tm_shape(tornado_sf_ga)+
  tm_dots(size=0.1)
```

The leaflet library is used in lots of web applications. The R version takes a little more work, but you can customize things like your basemap.

```{r}
library(leaflet)
pal <- colorNumeric(
  palette = "Blues",
  domain = tornado_ga$total)

leaflet(tornado_ga) %>%
  addProviderTiles(providers$Stamen.TonerHybrid) %>%
  addPolygons(stroke=FALSE,fillOpacity=0.8,color = ~pal(total))
```

There are other good mapping packages available. Mapview (https://r-spatial.github.io/mapview/articles/articles/mapview_01-basics.html) provides quick interactive maps, but at this moment only works on R version 3.5 and higher. The new cartography package (https://github.com/riatelab/cartography) also has helpful looking tools for mapping.

Learn more about spatial analysis in R in Manuel Gimond's web textbook: https://mgimond.github.io/Spatial/