---
title: "Geog6300: Univariate regression"
output: html_notebook
editor_options: 
  chunk_output_type: console
---

```{r setup}
library(tidyverse)
library(sf)
```


Here's our garden data from class
```{r}
med.age<-c(43, 21, 25, 42, 56, 59)
garden<-c(99, 65, 79, 75, 87, 81)
data<-data.frame(cbind(med.age, garden))
```

Now we can create a linear model. The actual model doesn't tell you much, but a summary of it does.
```{r}
lm(garden~med.age, data=data) 
model<-lm(garden~med.age, data=data)
summary(model)
```

You can also plot this model. The abline function plots a regression line created by the model.
```{r}
plot(garden~med.age, data=data)
abline(model) 
```

Let's do a more complex model: pollen data. Does yearly mean temperature predict the levels of pollen from birch trees (Betula)? We will load the data from Github and then make it spatial using the st_as_sf function in the sf package.

```{r}
Midwest_Pollen_Data<-read_csv("data/Midwest_Pollen_Data.csv")
pollen_data_sf<-st_as_sf(Midwest_Pollen_Data,coords=c(x="longitud",y="latitude"),crs=4326)
```

Let's look at the distribution of these variables first.
```{r}
hist(pollen_data_sf$Betula)
qqnorm(pollen_data_sf$Betula);qqline(pollen_data_sf$Betula)

hist(pollen_data_sf$tmeanyr)
qqnorm(pollen_data_sf$tmeanyr);qqline(pollen_data_sf$tmeanyr)

ggplot(pollen_data_sf,aes(y=Betula,x=tmeanyr))+
  geom_point()
```

We can use tmap to visualize these points.

```{r}
library(tmap)
tmap_mode("view")
tm_shape(pollen_data_sf) +
  tm_dots("Betula",size=.2)

tm_shape(pollen_data_sf) +
  tm_dots("tmeanyr",size=.2)
```

Now let's create a model.
```{r}
model<-lm(Betula~tmeanyr,data=pollen_data_sf)
summary(model)
```

We can plot out the model.
```{r}
plot(Betula~tmeanyr,data=pollen_data_sf)
abline(model)
```

##There's several diagnostics that can be used for regression.

*Normality of residuals: Plot/test the residuals. The "residuals" function pulls residuals from the model.
```{r}
pollen_data_sf$residuals<-residuals(model) #Pull the residuals from the model
hist(pollen_data_sf$residuals)
qqnorm(pollen_data_sf$residuals)
shapiro.test(pollen_data_sf$residuals)
```

Map the residuals using tmap
```{r}
tm_shape(pollen_data_sf) +
  tm_dots("residuals",size=.2)
```

*Heteroskedasticity: We can use the Breusch-Pagan test in the lmtest package. The null hypothesis is that the data is uniform, NOT heteroskedastic.
```{r}
library(lmtest)
bptest(model)
```

*Outliers: We'll use Cook's Distance to assess outliers This is adapted from: http://r-statistics.co/Outlier-Treatment-With-R.html See also https://onlinecourses.science.psu.edu/stat501/node/340 and http://www.statisticshowto.com/cooks-distance/.

In th example below, we plot the Cooke's distance for each observation, which shows the leverage it has in the model as a whole. We then add a cutoff line that shows values four times greater than the mean. Lastly, we add labels that give the row number of each identified outlier.

```{r}
cooks_dist<-cooks.distance(model)

plot(cooks_dist, pch="*", cex=2, main="Influential Obs by Cooks distance")
abline(h = 4*mean(cooks_dist, na.rm=T), col="red") 
text(x=1:length(cooks_dist)+1, y=cooks_dist, labels=ifelse(cooks_dist>4*mean(cooks_dist, na.rm=T),names(cooks_dist),""), col="red")  
```

We can go even further. Let's join these distances back to the dataset. We can add a "dummy variable" using if_else highlighting just those observations above that cutoff line.

```{r}
cooks_d<-data.frame(cooks_dist)
cutoff<-mean(cooks_dist)*4

pollen_data_sf<-pollen_data_sf %>%
  bind_cols(cooks_d) %>%
  mutate(outlier=if_else(cooks_dist>cutoff,1,0)) 
```

Where do those outliers fit in the overall distribution?

```{r}
ggplot(pollen_data_sf,aes(y=Betula,x=tmeanyr,color=outlier))+
  geom_point()
```

Where are these outliers?
```{r}
tm_shape(pollen_data_sf) +
  tm_dots("outlier",size=0.2)
```

Let's run a model without the outliers.
```{r}
pollen_data_sf1<-pollen_data_sf %>%
  filter(outlier==0)
model_rev<-lm(Betula~tmeanyr,data=pollen_data_sf1)
summary(model_rev)

plot(Betula~tmeanyr,data=pollen_data_sf1)
abline(model_rev)
```

##You try it!
Run a model for one other tree species. Assess the model results and run diagnostics.