---
title: "R Notebook"
output: html_notebook
editor_options: 
  chunk_output_type: console
---
```{r setup}
library(tidyverse)
```


##Post-hoc ANOVA analysis

Let's start by loading our in class "data"
```{r}
part_id<-c(4,5,6,13,14,15,8,9,16,17,18,7,1,2,3,10,11,12)
drug<-c("anxifree","anxifree","anxifree","anxifree","anxifree","anxifree","joyzepam","joyzepam","joyzepam","joyzepam","joyzepam","joyzepam","placebo","placebo","placebo","placebo","placebo","placebo")
therapy<-c("no.therapy","no.therapy","no.therapy","CBT","CBT","CBT","no.therapy","no.therapy","CBT","CBT","CBT","no.therapy","no.therapy","no.therapy","no.therapy","CBT","CBT","CBT")
result<-data.frame(c(0.6,0.4,0.2,1.1,0.8,1.2,1.7,1.3,1.8,1.3,1.4,1.4,0.5,0.3,0.1,0.6,0.9,0.3))
anova_data_navarro<-data.frame(cbind(part_id,drug,therapy,result))
names(anova_data_navarro)<-c("part_id","drug","therapy","result")
```

Next we create an ANOVA object and use summary to look at our results. What's our null hypothesis mean? How do we interpret a p<0.05? What other information is provided in the output?
```{r}
drugtest<-aov(result~drug,anova_data_navarro)
summary(drugtest)
```

Based on our results we can run follow up t-tests.
```{r}
placebo<-subset(anova_data_navarro,drug=="placebo")[,4]
anxifree<-subset(anova_data_navarro,drug=="anxifree")[,4]
joyzepam<-subset(anova_data_navarro,drug=="joyzepam")[,4]
t.test(placebo,anxifree, var.equal=TRUE)
t.test(placebo,joyzepam, var.equal=TRUE)
t.test(anxifree,joyzepam, var.equal=TRUE)
```

We can also use pairwise t test without and with Bonferroni correction.
```{r}
pairwise.t.test(x=anova_data_navarro$result, 
                g=anova_data_navarro$drug,
                var.equal=TRUE,
                pool.sd=FALSE) 

pairwise.t.test(x=anova_data_navarro$result, 
                g=anova_data_navarro$drug,
                var.equal=TRUE,
                pool.sd=FALSE,
                p.adjust.method="bonferroni")
```


Tukey HSD test is a more sophisticated test with correction. We use the ANOVA object for it. How is this output similar or different to pairwise.t.test?
```{r}
TukeyHSD(drugtest)
```

Testing residuals for normality--just use the "residuals" function on the ANOVA object. We can then create a qqplot and use shapiro.test to look at normality.
```{r}
drugtest.resid<-residuals(drugtest)
qqnorm(drugtest.resid);qqline(drugtest.resid)
hist(drugtest.resid)
shapiro.test(drugtest.resid)
```

Running a Fligner-Killeen test checks for equal variances between groups.
```{r}
fligner.test(result~drug,data=anova_data_navarro)
```

You could also treat the drug trial data as non-parametric. kruskal.test is the non-parametric version.
```{r}
kruskal.test(anova_data_navarro$result~anova_data_navarro$drug)
```

In class: Test for differences in pollen levels for species across regions. Do any species NOT have regional differences? The region variable is Region_fac.
```{r}
pollen_data<-read_csv("https://github.com/jshannon75/geog4300/raw/master/data/Midwest_Pollen_Data.csv")
```



