---
title: "Geog6300: Chi square tests"
output: html_notebook
editor_options: 
  chunk_output_type: console
---

Chi-square tests. First, we create the data and then run the test. Here we use the Thanksgiving.chisquare dataset from lecture.
```{r}
white<-c(15, 25)
dark<-c(25, 10)
tofurkey<-c(3, 7)
thanksgiving<-data.frame(cbind(white, dark, tofurkey))

chisq.test(thanksgiving)
```

Here's the same procedure for our in class problem.

```{r}
car<-c(58, 199, 205)
transit<-c(21, 49, 32)
walk<-c(36, 64, 29)
survey<-data.frame(cbind(car, transit, walk))
chisq.test(survey)
```

In class challenge: Look at the "tornado_points" data from earlier in the semester

```{r}
library(tidyverse)
tornado<-read_csv("https://github.com/jshannon75/geog4300/raw/master/data/tornado_points.csv")
```

Is there a difference in the severity of tornados (MAG) across states (ST)?

You will have to create a table of counts for each MAG category by state using the table function or group_by/summarise and then spread. Then run the chi-squared test.

