---
title: "Geog6300: Correlation"
output: html_notebook
editor_options: 
  chunk_output_type: console
---

```{r setup}
library(tidyverse)
```


Correlation

Here's the fictitious dataset we used in lecture.
```{r}
MHI<-c(40000, 59000, 61000, 66000, 68000, 71000, 78000, 82000, 89000, 103000, 104000,113000, 129000, 141000,156000)
VCR<-c(45.2, 44.6, 39.2, 42.4, 37.4, 29.4, 38.1, 28.5, 21.4, 19.4, 25.1, 28.5, 22.5, 17.3, 13.5)

crimedata<-data.frame(cbind(MHI,VCR))
```


A simple scatterpot shows possible correlation.

```{r}
plot(crimedata$MHI, crimedata$VCR)

#Same plot in ggplot
ggplot(crimedata,aes(x=MHI,y=VCR))+geom_point()
```

Use the cor function to see the correlation.

```{r}
#As separate vectors
cor(MHI, VCR)

#As columns in a data frame
cor(crimedata$MHI,crimedata$VCR)
```


Correlations are often shown as a matrix. Here, we can use the pollen data for the Midwest. The rcorr function in the Hmisc package does this well.

```{r}
require(Hmisc)
Midwest_Pollen_Data<-read_csv("https://github.com/jshannon75/geog4300/raw/master/data/Midwest_Pollen_Data.csv")
pollen<-Midwest_Pollen_Data[,c(6:15,21)] #Take just the first 10 species
names(pollen)
rcorr(as.matrix(pollen))#The data must be formatted as a matrix
```

Note that this shows both the correlation coefficient and its p value.

You can visualize the relationship using the pairs function, which is part of "base R" (comes installed). This creates scatterplots for every variable pair
```{r}
pairs(pollen)
```

There's lots of ways to visualize correlation matrices, and we often want to look at lots of variable combinations at once. Here's ggcorrplot. (More info: http://www.sthda.com/english/wiki/ggcorrplot-visualization-of-a-correlation-matrix-using-ggplot2)

```{r}
require(ggcorrplot)
correlation_matrix <- cor(pollen)
ggcorrplot(correlation_matrix, 
           type = "upper", 
           method="circle")

ggcorrplot(correlation_matrix,
    lab=TRUE, #Include numbers
    type = "lower", 
    insig = "blank") #Leave out insignificant cells
```

The corrgram function in the corrgram package is another option. https://rawgit.com/kwstat/corrgram/master/vignettes/corrgram_examples.html

```{r}
require(corrgram)
corrgram(correlation_matrix, order=FALSE,
         lower.panel=panel.conf,
         upper.panel=panel.shade, text.panel=panel.txt,
         main="Census correlations")
```


For non-parametric data, specify spearman's rho in the rcorr package.
```{r}
rcorr(as.matrix(pollen), type="spearman")
```

You can also specify test for the cor command.
```{r}
M<-cor(pollen.pct,method="spearman")
corrgram(M, panel=panel.shade)
```



