---
title: "Geog6300: Chi square tests"
output: html_notebook
editor_options: 
  chunk_output_type: console
---

Chi-square tests. First, we create the data and then run the test. Here we use the Thanksgiving.chisquare dataset from lecture.
```{r}
white<-c(15, 25)
dark<-c(25, 10)
tofurkey<-c(3, 7)
thanksgiving<-data.frame(cbind(white, dark, tofurkey))

chisq.test(thanksgiving)
```

Here's the same procedure for our in class problem.

```{r}
car<-c(58, 199, 205)
transit<-c(21, 49, 32)
walk<-c(36, 64, 29)
survey<-data.frame(cbind(car, transit, walk))
chisq.test(survey)
```

Creating a contingency table is possible with tidyverse tools. Is there a difference in the severity of tornados (MAG) across states (ST)?  

```{r}
library(tidyverse)
tornado<-read_csv("https://github.com/jshannon75/geog4300/raw/master/data/tornado_pointsSE.csv")

tornado_table<-tornado %>%
  group_by(ST,MAG) %>%
  summarise(count=n()) %>%
  pivot_wider(names_from=MAG,values_from=count,values_fill=0) %>%
  select(-`-9`) %>%
  ungroup()
```

We can then do a chi-square on just the count variables, removing the state variable.

```{r}
chisq.test(tornado_table %>% select(-ST))
```

