---
title: "Kernel density and quadrats"
output: github_notebook
editor_options: 
  chunk_output_type: console
---

For this script, you'll need tidyverse and sf, as well as a new (to us) package called spatstat.

```{r}
library(tidyverse)
library(sf)

#install.packages("spatstat")
library(spatstat)
```

We will continue work with tornado data for this anlaysis.

```{r}
tornado_points<-read_csv("https://github.com/jshannon75/geog4300/raw/master/data/tornado_points.csv")
```

Quadrat analysis in R often uses a package called spatstat. A lot of info on spatstat is available here: https://research.csiro.au/software/wp-content/uploads/sites/6/2015/02/Rspatialcourse_CMIS_PDF-Standard.pdf. With spatstat, you can create square quadrants of varying sizes

To use the spatstat package, you must first determine the maximum and minimum values for your lat/long coordinates. Here, we use the min and max functions to do so.

```{r}
maxlat<-max(tornado_points$CLAT)
minlat<-min(tornado_points$CLAT)
maxlon<-max(tornado_points$CLONG)
minlon<-min(tornado_points$CLONG)
```

Spatstat needs a "ppp" object. To create that, use the ppp function, providing X and Y points plus the range of your X and Y coordinates.

```{r}
tornado_ppp<-ppp(tornado_points$CLONG,tornado_points$CLAT,
                 c(minlon,maxlon),c(minlat,maxlat))
plot(tornado_ppp)
```

Spatstat can give us some basic statistics. Remember that the unit is in decimal degrees.
```{r}
summary(tornado_ppp) #Remember that the unit here is decimal degrees
```

We can use the density function to create a kernel density map.

```{r}
plot(density(tornado_ppp))
```

We can change the size of the "kernel" used to interpolate this map. Again, remember that the unit here is decimal degrees. Which is best?

```{r}
plot(density(tornado_ppp,0.1))
plot(density(tornado_ppp,0.25))
plot(density(tornado_ppp,0.5))
plot(density(tornado_ppp,1))
```

Spatstat can also create a contour map. The lines indicate the value of the kernel density map.

```{r}
contour(density(tornado_ppp))
```

Spatstat can create a quadrat map with square cells. The two numbers are the count of columns and rows.

```{r}
plot(quadratcount(tornado_ppp,10,5)) 
plot(quadratcount(tornado_ppp,15,10)) 
plot(quadratcount(tornado_ppp,6,3)) 
```

You can combine these maps with the add=TRUE parameter.
```{r}
kernel_map<-density(tornado_ppp,0.5)
quad_map<-quadratcount(tornado_ppp,15,10)

plot(kernel_map)
plot(quad_map, add=TRUE,col="white")

```

You try it! Create a kernel density map for just tornadoes in Georgia
In your lab: point pattern anlaysis of crimes in Spokane, WA.